---
title: "AMOVA for incomplete genomic data"
author: "Thierry Gosselin"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: amova_references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{AMOVA for incomplete genomic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(assigner)
```

# Purpose

Analysis of molecular variance (AMOVA) partitions molecular variation among
nested biological levels using a matrix of squared distances
[@Excoffier1992]. It is a general framework rather than a single genetic
distance or a single statistic.

Several mature programs and R packages already implement AMOVA. The goal of
`assigner::amova_genomic()` is not to replace or discredit them. Its narrower
goal is to make the observations contributing to AMOVA explicit when a genomic
matrix is incomplete: which individuals and populations contributed at each
locus, which loci were retained, whether a distance was Euclidean, and how the
global variance components were assembled.

This vignette explains the theory, compares existing implementations, and
documents the choices made in `assigner`.

# AMOVA in brief

## From observations to squared distances

For (n) molecular observations, let (Delta=(\delta_{ij}^2)) be the matrix
of squared distances. The total sum of squared deviations can be written as

\[
SS_T = \frac{1}{2n}\sum_{i=1}^{n}\sum_{j=1}^{n}\delta_{ij}^2.
\]

Equivalent sums are calculated within populations and, when present, within
successively higher groups. Differences between these sums provide the sums of
squares assigned to each hierarchical level. Dividing by the corresponding
degrees of freedom gives mean squares. Method-of-moments equations then
estimate variance components while accounting for unequal sample sizes
[@Excoffier1992].

For a hierarchy `REGION/POP_ID`, the variance components are conventionally
written as

* (\sigma_a^2): among regions;
* (\sigma_b^2): among populations within regions;
* (\sigma_c^2): within populations.

The associated statistics are

\[
\Phi_{ST}=\frac{\sigma_a^2+\sigma_b^2}
                 {\sigma_a^2+\sigma_b^2+\sigma_c^2},
\]

\[
\Phi_{CT}=\frac{\sigma_a^2}
                 {\sigma_a^2+\sigma_b^2+\sigma_c^2},
\qquad
\Phi_{SC}=\frac{\sigma_b^2}{\sigma_b^2+\sigma_c^2}.
\]

The notation describes correlations of molecular differences at specified
hierarchical levels. It should not be confused automatically with every
estimator called (F_{ST}). Weir and Cockerham's estimator, for example, is
derived from allelic correlations and sampling theory rather than from an
arbitrary molecular-distance matrix [@WeirCockerham1984].

## The distance is part of the biological model

The original AMOVA paper emphasizes that alternative distance matrices encode
alternative assumptions about molecular evolution [@Excoffier1992]. Examples
include:

* identity distance: zero for identical states and one otherwise;
* nucleotide distance: the number or proportion of nucleotide substitutions;
* allele-sharing distances for codominant genotypes;
* Euclidean or Manhattan distances between allele dosages;
* user-defined distances appropriate to a marker system.

Changing the distance can change the estimand. Consequently, a function should
record the distance used rather than label all analyses simply as "AMOVA".

Classical AMOVA uses a squared Euclidean distance matrix. A non-Euclidean
dissimilarity can generate negative eigenvalues and variance components that
do not have the usual Euclidean decomposition. `assigner` therefore records an
eigenvalue diagnostic and can apply a Lingoes correction. A correction changes
the geometry and must be reported; it is not merely a numerical detail.

## Negative variance components

Sampling estimates of variance components can be negative. A negative estimate
does not necessarily indicate a programming error, and replacing it silently
with zero changes the estimator. `amova_genomic()` retains negative components.
Biological interpretation should consider sampling uncertainty, the distance,
and the sampling design.

# Standardized differentiation is a second calculation

(\Phi_{ST}) may have a low attainable maximum when within-population diversity
is high. Meirmans proposed standardizing the statistic by its maximum possible
value conditional on the observed within-population component
[@Meirmans2006]:

\[
\Phi'_{ST}=\frac{\Phi_{ST}}{\Phi_{ST(max)}}.
\]

The maximum must be defined by the distance model. It is not, in general, the
largest distance observed in a sample. Dividing every locus by its observed
maximum makes the result depend on sample composition and missingness.

Following the logic used by GenoDive and Stacks, `assigner` currently computes
the standardized statistic only with identity distance. The observed
within-population distances are retained, whereas between-population distances
are replaced by their theoretical maximum. Other distances can be used for
(\Phi_{ST}), but they do not receive an automatic (\Phi'_{ST}) unless a
defensible theoretical maximum is supplied.

This is why (\Phi_{ST}) and (\Phi'_{ST}) need not use the same distance
matrix. Reporting both can be useful because ordinary and standardized
differentiation answer related but distinct questions [@MeirmansHedrick2011].

# Why incomplete genomic data need special treatment

With hundreds or thousands of RADseq loci, requiring every individual to be
called at every locus can discard most of the dataset. At the other extreme,
replacing missing values with zero treats absence of information as a measured
allele count. A single pairwise distance calculated from a different set of
loci for every pair can also be difficult to audit and may not be Euclidean.

Missingness is additionally capable of being informative. Coverage, restriction
site polymorphism, mapping quality, population ancestry, and laboratory batch
can all affect whether a genotype is observed. No distance formula can make
such missingness ignorable by itself.

`assigner` therefore treats missing-data policy as part of the analysis:

1. **Locus-wise** (default): remove uncalled observations only for the current
   locus, then recompute the represented hierarchy, sample sizes, degrees of
   freedom, and unequal-sample-size coefficients.
2. **Filter**: apply explicit call-rate and representation thresholds, then use
   the locus-wise calculation.
3. **Complete**: retain only individuals called at all retained loci. This is
   provided as a diagnostic or for genuinely dense datasets.

No imputation is performed. Imputation belongs in `grur`, where multiple
completed datasets and their uncertainty can be generated explicitly. A later
`assigner` interface can combine AMOVA estimates across those datasets rather
than hiding imputation inside the variance calculation.

# Existing implementations

The following comparison describes differences in scope and interface. It is
not a ranking of scientific quality.

## `ade4`

`ade4::amova()` is a direct and general implementation based on haplotype-count
tables, an optional distance object, and optional nested structures. It verifies
that a supplied distance is Euclidean and `ade4::randtest()` provides
randomization tests. `ade4` is a foundational multivariate-analysis package
developed by an expert group that includes Stéphane Dray, Anne-Béatrice Dufour,
Jean Thioulouse, Daniel Chessel, Thibaut Jombart, and contributors including
Pierre Legendre [@DrayDufour2007].

Its AMOVA interface expects the user to provide valid sample counts and a
complete distance object. That is a sensible separation of responsibilities,
but it means that locus-dependent missingness must be resolved before calling
the function. It does not itself maintain an audit of different contributing
individuals and populations for thousands of genomic loci.

## `pegas`

`pegas::amova()` uses a convenient formula interface such as
`distance ~ region/population`. It accepts a distance matrix, calculates
hierarchical variance components, and implements level-specific permutations
[@Paradis2010]. Its source uses the exact Excoffier coefficients for one- and
two-level hierarchies and documented approximate formulae for deeper
hierarchies.

As with `ade4`, the distance matrix is the analytical input. Missing values in
that matrix are warned about, but constructing a defensible distance matrix
from sparse genomic observations remains the user's responsibility.

## `poppr`

`poppr::poppr.amova()` provides an important bridge from `adegenet` genetic
objects to either the `ade4` or `pegas` AMOVA engines. It supports clonal data,
haplotype splitting, mixed reproductive systems, missing-data preprocessing,
and corrections for non-Euclidean distances [@Kamvar2014; @Kamvar2015].

For `genind` objects, missing data are handled before AMOVA through options such
as removing loci or genotypes or replacing missing allele frequencies. For
`genlight`, its bitwise distance can scale pairwise comparisons for missing
observations. These are useful general policies, but they differ from rebuilding
the AMOVA sampling coefficients separately from the observations present at
every locus. `poppr` also has a broader and different purpose: population
genetics with particular strength for clonal and partially clonal organisms.

## GenoDive

GenoDive implements hierarchical AMOVA, ordinary and standardized
differentiation, numerous distances, filtering, and simple procedures for
filling missing genotypes [@Meirmans2020]. The published description lists
three procedures: randomly drawing alleles from overall allele frequencies,
randomly drawing alleles from population allele frequencies, and restoring
polyploid dosage from incomplete genotypes and population allele frequencies.
These are more specific than replacing every missing value by a column mean,
but they are single-completion, frequency-based approaches. They should not be
described as a modern multiple-imputation framework designed to propagate
uncertainty from the high and structured missingness often observed in RADseq
datasets. GenoDive's standardized AMOVA is closely connected to Meirmans'
derivation of (\Phi'_{ST}) [@Meirmans2006], making it an important reference
implementation and a useful independent comparison for `assigner`.

GenoDive is a standalone application rather than an R package with an exposed R
implementation. The published papers document its capabilities, but not every
internal computational decision can be verified from the material distributed
with `assigner`. Comparisons should therefore use controlled input datasets and
document software versions and settings rather than infer undocumented details.

## Stacks

The `populations` program in Stacks was designed around RADseq data
[@RochetteCatchen2019]. Its source illustrates a useful genomic strategy:
observed haplotypes are tabulated independently at each locus, populations
without valid observations are excluded at that locus, and the AMOVA sample-size
coefficients are recalculated from the represented haplotypes. It supports
minimum sample proportions per population and minimum numbers of represented
populations.

Stacks calculates haplotype (\Phi_{ST}) using nucleotide-substitution
distances. For standardized differentiation, it separately uses identity
distances and constructs a maximum between-population distance. These choices
strongly influenced the `assigner` design.

Stacks is nevertheless a complete RADseq processing pipeline with its own data
model and output conventions. It is not intended as a general R AMOVA engine
for arbitrary tidy genomic inputs or user-defined distances.

# What is different in `assigner`?

`assigner::amova_genomic()` combines established AMOVA calculations with an
explicit incomplete-genomic-data contract:

* long genomic data are analyzed locus by locus;
* the observations and hierarchical units contributing to every locus are
  recalculated rather than inferred from the full input strata;
* global statistics are ratios of **summed locus variance components**, not
  arithmetic means of locus (\Phi)-statistics;
* marker call rates, represented groups, sample sizes, retained loci,
  eigenvalues, and geometric corrections are returned for audit;
* distance methods have explicit names and allow a custom squared-distance
  function;
* ordinary (\Phi)-statistics and standardized (\Phi'_{ST}) are kept as
  separate calculations;
* permutation occurs at the exchangeable unit appropriate to the tested
  hierarchy level;
* imputation is intentionally outside the function.

The contribution is therefore not a new definition of AMOVA. It is a
reproducible way to connect AMOVA to incomplete next-generation genotype data
while exposing decisions that are otherwise commonly made during preprocessing.

# Running `amova_genomic()`

The public input follows the rest of `assigner`: `data` may be a GDS file path
or an open GDS object handled by `genometranslator::read_genome()`. Diploid
alternate-allele dosage is detected automatically from standard GDS genotype
fields. A long data frame remains supported for programmatic use and examples.
Hierarchy columns are ordered from highest to lowest and may come from GDS
metadata or the separate `strata` argument.

```{r example-data}
set.seed(2026)

example_amova <- expand.grid(
  INDIVIDUALS = sprintf("ind_%02d", 1:24),
  MARKERS = paste0("loc_", 1:12),
  stringsAsFactors = FALSE
)

individual_strata <- data.frame(
  INDIVIDUALS = sprintf("ind_%02d", 1:24),
  REGION = rep(c("north", "south"), each = 12),
  POP_ID = rep(c("n1", "n2", "s1", "s2"), each = 6)
)

example_amova <- merge(example_amova, individual_strata, by = "INDIVIDUALS")
population_effect <- c(n1 = 0.1, n2 = 0.6, s1 = 1.2, s2 = 1.7)
example_amova$GT_BIN <- rbinom(
  nrow(example_amova), 2,
  plogis(-1 + population_effect[example_amova$POP_ID])
)
example_amova$GT_BIN[sample(nrow(example_amova), 30)] <- NA

head(example_amova)
```

## Locus-wise AMOVA

```{r locuswise}
fit <- amova_genomic(
  data = example_amova,
  hierarchy = c("REGION", "POP_ID"),
  value = "GT_BIN",
  distance = "euclidean",
  missing = "locuswise",
  min.groups = 2,
  min.individuals = 2,
  standardized = FALSE
)

fit
head(fit$per_locus)
head(fit$marker_audit)
```

## Filtering before locus-wise calculation

```{r filtered}
fit_filtered <- amova_genomic(
  data = example_amova,
  hierarchy = c("REGION", "POP_ID"),
  value = "GT_BIN",
  distance = "euclidean",
  missing = "filter",
  min.call.rate = 0.75,
  min.groups = 4,
  min.individuals = 2,
  standardized = FALSE
)

fit_filtered$global
```

## Standardized differentiation

For a Meirmans-style standardized statistic, use an identity state and identity
distance. The `standardized` column is populated for `PHI_ST`; it is not
silently generalized to hierarchy-specific standardized statistics whose
maxima have not been implemented.

```{r standardized-example, eval = FALSE}
fit_identity <- amova_genomic(
  data = haplotype_data,
  hierarchy = c("REGION", "POP_ID"),
  value = "HAPLOTYPE",
  distance = "identity",
  missing = "locuswise",
  standardized = TRUE
)
```

## Hierarchical permutations

```{r permutations, eval = FALSE}
fit_permuted <- amova_genomic(
  data = example_amova,
  hierarchy = c("REGION", "POP_ID"),
  value = "GT_BIN",
  distance = "euclidean",
  missing = "locuswise",
  standardized = FALSE,
  permutations = 999,
  seed = 2026
)

fit_permuted$permutation$p_value
fit_permuted$permutation$design
```

### The finite permutation space is part of the result

Fitzpatrick's key observation is combinatorial, not a competing estimator of
AMOVA [@Fitzpatrick2009]. To test a group-level component, populations are the
exchangeable replicates. More individuals or loci can improve estimation of
within-population genetic variation, but they do not create more independent
ways to allocate the sampled populations among groups.

For group sizes \(k_1,\ldots,k_g\), `amova_genomic()` calculates the number of
unique allocations from the multinomial coefficient and removes duplicate
partitions caused by groups of equal size. For two groups containing two
populations each, there are only three unique partitions. Consequently the
smallest exact probability is \(1/3\), and a group-level test cannot attain
\(\alpha=0.05\), regardless of how many SNPs or individuals were sampled.

The `design` table reports, for every tested component:

* the unit that is exchangeable under its null hypothesis;
* the number of those units in the sampling design;
* the number of unique allocations;
* the theoretical minimum attainable \(P\)-value; and
* whether the requested `alpha` can be attained.

The function warns before interpretation when a test cannot attain `alpha`.
This is intentionally different from merely reporting the Monte Carlo
resolution \(1/(B+1)\) for `B` sampled permutations [@PhipsonSmyth2010]. The returned `p_value` is
not allowed below the exact finite-space bound; `monte_carlo_p` is retained for
auditing the random approximation.

### Genomic loci do not multiply the high-level replication

With incomplete genomic data, each locus can contain a different subset of
individuals. Nevertheless, one hierarchy randomization is drawn per component
and iteration and then applied consistently to every retained locus. Permuting
the hierarchy independently at every locus would incorrectly turn loci into
additional population-level replication and could make a weak sampling design
look much more informative than it is.

Fitzpatrick's conclusion has been carried into later sampling-design guidance
[@Werth2011]:
for a hierarchical group test, sampling additional populations per group is
more useful than adding individuals to a few populations. The limitation is
also a standard property of restricted randomization tests. It is therefore
appropriate to implement it as an audit and validity safeguard, not to present
it as a new form of AMOVA.

The AMOVA implementations reviewed for this vignette provide user-selected
randomization counts, but their ordinary result objects do not generally expose
the design-specific number of unique high-level allocations or warn that the
chosen significance level is unattainable. This does **not** mean their AMOVA
estimators are wrong, nor that their authors rejected Fitzpatrick's argument.
The distinction in `assigner` is that the limitation is calculated and reported
alongside the component test, which is particularly important when thousands
of loci can give a misleading impression of replication.

### P-values are not effect-size uncertainty

Permutation tests and confidence intervals answer different questions. A
permutation \(P\)-value asks how compatible the observed component is with a
particular null randomization. It does not show how precisely a \(\Phi\)-statistic
was estimated, and a small value does not imply biologically important
differentiation. Conversely, a scientifically meaningful estimate can remain
uncertain or have an unattainable permutation threshold when few populations
were sampled.

Confidence intervals for \(F\)- and \(\Phi\)-statistics are consequently useful:
they display uncertainty around the estimated magnitude and allow readers to
judge whether zero and biologically negligible effects remain plausible. They
should be reported alongside the estimate rather than treated merely as another
route to a significance label.

For hierarchical genomic AMOVA, however, the bootstrap design is part of the
estimand:

* resampling individuals within populations represents uncertainty from the
  sampled individuals, but cannot create population-level replication;
* resampling populations within their parent groups can address a higher-level
  component when populations are the intended sampling units;
* resampling single SNPs as if independent is anticonservative under linkage;
  a locus bootstrap is appropriate only for effectively independent loci, while
  a genomic block bootstrap requires genomic positions and defensible blocks;
* uncertainty caused by imputation or non-random missingness is a separate
  layer and should eventually be propagated from `grur` completed datasets.

`amova_genomic()` therefore offers explicitly labelled locus and genomic-block
percentile intervals. Locus resampling assumes effectively independent loci;
block resampling is preferred when linkage can be represented by an existing
block column or by chromosome, position, and a defensible physical block width.

```{r uncertainty, eval = FALSE}
fit_uncertainty <- amova_genomic(
  data = gds,
  hierarchy = c("REGION", "STRATA"),
  resampling = "block",
  chromosome = "CHROM", position = "POSITION", block.size = 1e6,
  bootstrap = 999, confidence = 0.95,
  population.jackknife = TRUE, seed = 2026
)
fit_uncertainty$uncertainty$report
fit_uncertainty$uncertainty$marker$intervals
fit_uncertainty$uncertainty$population.jackknife
```

The population jackknife refits AMOVA after omitting each lowest-level
population. It measures leverage, not population-sampling uncertainty. Marker
intervals likewise do not represent population replication.

# Validation strategy

Before treating this implementation as a reference estimator, it should be
validated in layers:

1. **Complete balanced data:** compare sums of squares, mean squares, variance
   components, and (\Phi)-statistics with `ade4`, `pegas`, GenoDive, and
   Stacks where their input models overlap.
2. **Complete unbalanced data:** verify unequal-sample-size coefficients against
   worked calculations and independent software.
3. **Incomplete data under MCAR:** introduce known call rates and compare bias,
   variance, coverage, and locus retention.
4. **Population- and genotype-associated missingness:** demonstrate where a
   locus-wise analysis remains biased because the missingness mechanism is not
   ignorable.
5. **Distance sensitivity:** compare identity, nucleotide, allele-sharing, and
   dosage distances without assuming they estimate the same biological
   quantity.
6. **Permutation calibration:** verify type-I error and power for each hierarchy
   level, with special attention to the number of exchangeable populations and
   groups.
7. **Performance:** profile realistic RADseq datasets before moving numerical
   kernels to C++. Statistical orchestration should remain visible in R.

# Recommended reading

The following sequence gives a useful path through the subject:

1. Excoffier, Smouse, and Quattro for the original distance-based AMOVA
   framework [@Excoffier1992].
2. Weir and Cockerham for sampling-based (F)-statistics and why estimators with
   similar labels need not be identical [@WeirCockerham1984].
3. Meirmans for standardized AMOVA differentiation [@Meirmans2006].
4. Meirmans and Hedrick for the interpretation of (F_{ST}), standardized
   (F'_{ST}), and related measures [@MeirmansHedrick2011].
5. Fitzpatrick for the limits of hierarchical permutation power
   [@Fitzpatrick2009].
6. The `ade4`, `pegas`, and `poppr` papers for their respective R ecosystems
   [@DrayDufour2007; @Paradis2010; @Kamvar2014; @Kamvar2015].
7. The GenoDive and Stacks 2 papers for practical population-genomic workflows
   [@Meirmans2020; @RochetteCatchen2019].

# Reproducibility checklist

An AMOVA report should state at least:

* software and version;
* hierarchy and sampling unit at every level;
* molecular distance and whether it was squared;
* missing-data and filtering policy;
* number of retained loci and samples per level;
* Euclidean diagnostic and any correction;
* whether global values were obtained from summed components or averaged locus
  statistics;
* definition of the maximum used for a standardized statistic;
* permutation unit, restrictions, seed, and number of permutations;
* whether negative variance components were retained or truncated.
