---
title: "Running assigner::assignment_ngs with several whitelists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running assigner::assignment_ngs with several whitelists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Prepare working environment

```{r Prepare working environment, eval = FALSE}
rm(list = ls())
if (!require("pak")) install.packages("pak")
pak::pkg_install("thierrygosselin/assigner")
library(assigner)
assigner::install_gsi_sim(fromSource = TRUE)
```

# Prepare whitelists

* Gather your whitelist in a single folder 
* Get the list of these file in a string with the command below:


```{r get whitelists, eval = FALSE}
all.whitelists <- list.files(path = "~/Desktop/whitelists_salmon", pattern = "whitelist")
all.whitelists # to see if the whitelists are all there
```

# Create a function to run your assignment

* Here the goal is to have a the argument you need selected
* The only argument that will change will be the whitelist
* you could have multiple arguments changing at the same time, but for simplicity, let's start with just one

```{r tailored assignment function, eval = FALSE}
whitelists_assigner <- function(x) {
  res <- assigner::assignment_ngs(
    data = "subset_melville_salmon.tsv", #change for what you want
    whitelist.markers = x, # don't change this one,
    assignment.analysis = "adegenet", #change for what you want
    common.markers = TRUE,#change for what you want
    marker.number = "all",# I think you get the idea...
    pop.levels = c(1,2,3,4,5,6,7,8,9,10,11),
    sampling.method = "random",
    iteration.method = 1,
    keep.gsi.files = FALSE
  )
}
```


# Run the assignment

* The trick here is to **map** these whitelists to your new function `whitelists_assigner`
* The results are in a list easily managed by R
* The `purrr` package is awesome for this, it's very similar to the `lapply` function, only better.
* If you have access to a MAC or a LINUX cluster, you could just replace the `purrr::map` function to `parallel::mclapply`

```{r run the assignment, eval = FALSE}
salmon.assignment.res <- purrr::map(
.x = all.whitelists, # your whitelist created above
.f = whitelists_assigner # your new function
)
```


